Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-07-31 Manuscript/Manuscript.tex

We therefore implement variable bedrock Ca+Mg concentration into GEOCLIM (\SI). The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) \cite{Hartmann2012a} and is represented by 6 categories: metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment. Each land pixel is assigned these lithologic categories at a resolution of 0.1\degrees $\times$ 0.1\degrees. The Ca+Mg concentrations of felsic, intermediate, and mafic lithologies are assigned based on the mean of data of these lithologic categories compiled in EarthChem (\url{www.earthchem.org/portal}). Given that GLiM does not distinguish ultramafic lithologies, such rocks are grouped with mafic rocks. As a result, the Ca+Mg concentration is likely an underestimate in regions of obducted ophiolites, such that the estimated effect of these regions on changing steady-state \pCOtwo is conservative. The weathering of carbonate does not contribute to long-term \COtwo consumption and its Ca+Mg concentration is ignored. The Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies are more difficult to define, since their chemical composition is strongly dependent on protolith composition and, in the case of siliciclastic sediment, the degree of previous chemical depletion. We explore a range of feasible Ca+Mg concentrations for metamorphic rocks and siliciclastic sediment during calibration of the silicate weathering component of GEOCLIM.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, erosion, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic lithologies to vary within reasonable bounds (\SI; Table S2). This approach results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 573 parameter combinations remain (\SI; Fig. S2). The resulting global \COtwo consumption of these filtered model runs all overlap with independently derived estimates of the global \COtwo degassing flux \cite{Gerlach2011a}, as they should given that the long-term carbon cycle is in steady state (\SI; Fig. S2).

@@

To determine the position of paleoshorelines in the SEAI over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the Australian Plate \cite{Harris2006a}. Most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. We also include changes in areas of presently submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S5). These tectonic drivers and others throughout the region led to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of the SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a climate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (\SI; Fig. S8). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

@@

Using the 573 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 526--678~ppm for 15~Ma, 457--516~ppm for 10~Ma, and 391--434~ppm \pCOtwo for 5~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of the SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$526--678~ppm. We note that these ``paleo-SEAI'' scenarios do not account for Neogene changes outside of the SEAI (e.g. changes in ocean/atmosphere circulation, volcanic degassing, and weathering fluxes elsewhere on Earth, discussed in \textit{Alternative Mechanisms for Neogene Cooling}). Therefore, these results do not imply that \pCOtwo would have been observed to be 526--678~ppm at 15~Ma, but rather that emergence of the SEAI alone is responsible for decreasing \pCOtwo by 240--392~ppm (526--678~ppm minus 286~ppm) over the past 15~m.y..

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation (\SI; Fig. S9). The \pCOtwo values from the 5~Ma ``paleo-SEAI'' scenario overlap with many proxy-based estimates \cite{Foster2017a} as well as values that emerge from approaches that assimilate climate and ice sheet model output with benthic \dO data \cite{van-de-Wal2011a, Berends2020a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early to mid-Miocene, indicating that the increase in subaerially exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo associated with emergence of the SEAI are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

Our model framework provides an opportunity to estimate the decrease in volcanic degassing flux necessary to achieve the same change in \pCOtwo predicted for the increase in global weatherability associated with the emergence of the SEAI over the past 15~m.y. If we use the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S3), GEOCLIM estimates a pre-industrial volcanic degassing flux of 4.1$\times$10$^{12}$~mol/yr to balance the silicate weathering flux at 286~ppm. If we then assume that this volcanic degassing flux did not change over the past 15~m.y., then GEOCLIM estimates that the increase in global weatherability associated with the emergence of the SEAI led to a change in \pCOtwo of $\sim$280~ppm (``increase in weatherability only'' scenario in Fig. \ref{fig:weatherability_curves}). If we instead assume that global weatherability did not change over the past 15~m.y., then we estimate that the volcanic degassing flux needs to have been $\sim$13\% greater at 15~Ma relative to the pre-industrial to drive the same $\sim$280~ppm change in \pCOtwo (``decrease in degassing only'' scenario in Fig. \ref{fig:weatherability_curves}). This $\sim$13\% value is higher than 10\%, the highest current estimate for the volcanic degassing flux at 15~Ma relative to the present day \cite{Van-Der-Meer2014a}.

However, changes in the volcanic degassing flux would have modulated changes in \pCOtwo associated with changes in global weatherability. For example, some proxy-based approaches as well as some model-data assimilation approaches estimate that mid-Miocene \pCOtwo was lower than 568~ppm (\SI; Fig. S9). Take a scenario in which \pCOtwo was 400~ppm at 15~Ma. If we assume that the \pCOtwo decrease to the pre-industrial value of 286~ppm was driven by the increase in global weatherability associated with emergence of the SEAI in conjunction with an increase in volcanic degassing which counteracts cooling by increasing the flux of \COtwo to the atmosphere (``increase in weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}), the volcanic degassing flux would have had to have been $\sim$7\% smaller than the pre-industrial. More robust constraints on \pCOtwo (\SI; Fig. S9) and/or volcanic degassing rates over the past 20~m.y. are needed to constrain which of the ``increase in weatherability only'' or ``increase in weatherability and degassing'' scenarios (Fig. \ref{fig:weatherability_curves}) is more representative of the mechanisms driving Neogene cooling.

@@

Alkalinity generated by the weathering of silicate Na and K consumes \COtwo until that alkalinity is removed via reverse weathering reactions \cite{Sun2016a}. Therefore, an imbalance in the Na and K cycles could consume significant quantities of \COtwo over multi-million year time-scales. However, the biogeochemical cycles of Na and K are currently poorly constrained, such that the residence time of these two cations, and therefore the time-scale over which an imbalance in their cycles could drive \COtwo consumption, is not well known. Estimates for these values in the literature range from XX to XX~m.y. We have therefore made the assumption in GEOCLIM that alkalinity generated by the weathering of silicate Na and K is instantaneously removed by reverse weathering reactions, but acknowledge that further work is necessary to evaluate whether such an assumption is valid.

@@

To create the 15~Ma ``paleo-SEAI'' curve shown in Figure \ref{fig:weatherability_curves}, we use the reconstructed paleoshorelines of the SEAI at 15~Ma (Fig. \ref{fig:shoreline_growth}A). We then select the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S3), and fix \pCOtwo at the 3 \pCOtwo levels at which the GFDL CM2.0 climate model experiments were computed (\SI). We then run GEOCLIM at each of these \pCOtwo levels until steady state is achieved (i.e. until the volcanic degassing flux is equal to the silicate weathering flux). We then repeat this process for the 10 and 5~Ma ``paleo-SEAI'' paleoshorelines and the present day shorelines to generate the 3 other weatherability curves. Each estimated \pCOtwo in Figure \ref{fig:scenario_pCO2} is the result of underlying weatherability curves that change with the different chemical weathering parameters.

>>>>

We therefore implement variable bedrock Ca+Mg concentration into GEOCLIM (\SI). The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) \cite{Hartmann2012a} and is represented by 6 categories: metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment. Each land pixel is assigned these lithologic categories at a resolution of 0.1\degrees $\times$ 0.1\degrees. The Ca+Mg concentrations of felsic, intermediate, and mafic lithologies are assigned based on the mean of data of these lithologic categories compiled in EarthChem (\url{www.earthchem.org/portal}). Given that GLiM does not distinguish ultramafic lithologies, such rocks are grouped with mafic rocks. As a result, the Ca+Mg concentration is likely an underestimate in regions of obducted ophiolites, such that the estimated effect of these regions on changing steady-state \pCOtwo is conservative \cite{Schopka2011a}. The weathering of carbonate does not contribute to long-term \COtwo consumption and its Ca+Mg concentration is ignored. The Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies are more difficult to define, since their chemical composition is strongly dependent on protolith composition and, in the case of siliciclastic sediment, the degree of previous chemical depletion. The siliciclastic sediment lithologic category also includes sedimentary sequences in which any carbonate is identified but is not the dominant lithology \cite{Hartmann2012a}. We explore a range of feasible Ca+Mg concentrations for metamorphic rocks and siliciclastic sediment during calibration of the silicate weathering component of GEOCLIM, but assume that all pixels of a given lithologic category share the same Ca+Mg concentration.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, erosion, and regolith thickness are poorly constrained. Rather than prescribing single values, we test several reasonable values for each of these four parameters along with the Ca+Mg concentration of metamorphic and siliciclastic lithologies (\SI; Table S2). We then permute all possible combinations of these reasonable values for the six parameters, leading to 93,600 unique parameter combinations (i.e. permutations). For each combination, we compute spatially resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 573 parameter combinations remain (\SI; Fig. S3). The resulting global \COtwo consumption of these filtered model runs all overlap with independently derived estimates of the global \COtwo degassing flux \cite{Gerlach2011a}, as they should given that the long-term carbon cycle is in steady state (\SI; Fig. S3).

@@

To determine the position of paleoshorelines in the SEAI over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the Australian Plate \cite{Harris2006a}. Most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. We also include changes in areas of presently submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S6). These tectonic drivers and others throughout the region led to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of the SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a climate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (\SI; Fig. S9). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

@@

Using the 573 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 526--678~ppm for 15~Ma, 457--516~ppm for 10~Ma, and 391--434~ppm \pCOtwo for 5~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of the SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$526--678~ppm. Increasing seawater Mg/Ca since ca. 15~Ma \cite{Higgins2012a} is consistent with an increasing proportion of the global silicate weathering flux being derived from mafic and ultramafic sources. We note that these ``paleo-SEAI'' scenarios do not account for Neogene changes outside of the SEAI (e.g. changes in ocean/atmosphere circulation, volcanic degassing, and weathering fluxes elsewhere on Earth, discussed in \textit{Alternative Mechanisms for Neogene Cooling}). Therefore, these results do not imply that \pCOtwo would have been observed to be 526--678~ppm at 15~Ma, but rather that emergence of the SEAI alone is responsible for decreasing \pCOtwo by 240--392~ppm (526--678~ppm minus 286~ppm) over the past 15~m.y.

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation (\SI; Fig. S10). The \pCOtwo values from the 5~Ma ``paleo-SEAI'' scenario overlap with many proxy-based estimates \cite{Foster2017a} as well as values that emerge from approaches that assimilate climate and ice sheet model output with benthic \dO data \cite{van-de-Wal2011a, Berends2020a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early to mid-Miocene, indicating that the increase in subaerially exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo associated with emergence of the SEAI are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

@@

Our model framework provides an opportunity to estimate the decrease in volcanic degassing flux necessary to achieve the same change in \pCOtwo predicted for the increase in global weatherability associated with the emergence of the SEAI over the past 15~m.y. If we use the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S4), GEOCLIM estimates a pre-industrial volcanic degassing flux of 4.1$\times$10$^{12}$~mol/yr to balance the silicate weathering flux at 286~ppm. If we then assume that this volcanic degassing flux did not change over the past 15~m.y., then GEOCLIM estimates that the increase in global weatherability associated with the emergence of the SEAI led to a change in \pCOtwo of $\sim$280~ppm (``increase in weatherability only'' scenario in Fig. \ref{fig:weatherability_curves}). If we instead assume that global weatherability did not change over the past 15~m.y., then we estimate that the volcanic degassing flux needs to have been $\sim$13\% greater at 15~Ma relative to the pre-industrial to drive the same $\sim$280~ppm change in \pCOtwo (``decrease in degassing only'' scenario in Fig. \ref{fig:weatherability_curves}). This $\sim$13\% value is higher than 10\%, the highest current estimate for the volcanic degassing flux at 15~Ma relative to the present day \cite{Van-Der-Meer2014a}.

However, changes in the volcanic degassing flux would have modulated changes in \pCOtwo associated with changes in global weatherability. For example, some proxy-based approaches as well as some model-data assimilation approaches estimate that mid-Miocene \pCOtwo was lower than 568~ppm (\SI; Fig. S10). Take a scenario in which \pCOtwo was 400~ppm at 15~Ma. If we assume that the \pCOtwo decrease to the pre-industrial value of 286~ppm was driven by the increase in global weatherability associated with emergence of the SEAI in conjunction with an increase in volcanic degassing which counteracts cooling by increasing the flux of \COtwo to the atmosphere (``increase in weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}), the volcanic degassing flux would have had to have been $\sim$7\% smaller than the pre-industrial. More robust constraints on \pCOtwo (\SI; Fig. S10) and/or volcanic degassing rates over the past 20~m.y. are needed to constrain which of the ``increase in weatherability only'' or ``increase in weatherability and degassing'' scenarios (Fig. \ref{fig:weatherability_curves}) is more representative of the mechanisms driving Neogene cooling.

@@

Alkalinity generated by the weathering of silicate Na and K consumes \COtwo until that alkalinity is removed via reverse weathering reactions \cite{Sun2016a}. Therefore, sustained imbalances in the Na and K cycles could consume/generate significant quantities of \COtwo over multi-million year time-scales. However, the biogeochemical cycles of Na and K are currently poorly constrained. As such, the residence time of these two cations, and therefore the time-scale over which an imbalance in their cycles could drive \COtwo consumption/generation, is not well known. Estimates for these values in the literature range from $\sim$45-80~m.y. for Na, and $\sim$8-10~m.y. for K \cite{Emerson2008a, Lecuyer2016a}. Furthermore, it is not clear how to parameterize the output fluxes of Na and K. We have therefore made the assumption in GEOCLIM that alkalinity generated by the weathering of silicate Na and K is instantaneously removed by reverse weathering reactions, but acknowledge that further work is necessary to evaluate how important imbalances in the Na and K cycles are for the geologic carbon cycle.

@@

To create the 15~Ma ``paleo-SEAI'' curve shown in Figure \ref{fig:weatherability_curves}, we use the reconstructed paleoshorelines of the SEAI at 15~Ma (Fig. \ref{fig:shoreline_growth}A). We then select the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI; Fig. S4), and fix \pCOtwo at the 3 \pCOtwo levels at which the GFDL CM2.0 climate model experiments were computed (\SI). We then run GEOCLIM at each of these \pCOtwo levels until steady state is achieved (i.e. until the volcanic degassing flux is equal to the silicate weathering flux). We then repeat this process for the 10 and 5~Ma ``paleo-SEAI'' paleoshorelines and the present day shorelines to generate the 3 other weatherability curves. Each estimated \pCOtwo in Figure \ref{fig:scenario_pCO2} is the result of underlying weatherability curves that change with the different chemical weathering parameters.

### 2020-07-31 Manuscript/Supplementary_Information.tex

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/world_lithology.jpg}
    \caption{Distribution of lithologies at 0.1\degrees $\times$ 0.1\degrees resolution used in GEOCLIM modified from \cite{Hartmann2012a}.}
    \label{fig:world_lithology}
\end{figure}

@@

@@

Proxy-based \pCOtwo estimates for the past 20~m.y. are shown in Figure \ref{fig:pCO2_proxies} \citep{Bereiter2015a, Foster2017a, Ji2018a, Cui2020a}. Even when stringent quality criteria and the latest understanding of each of the \pCOtwo proxies have been applied to available \pCOtwo records, both significant uncertainty in the estimated \pCOtwo for any given data point as well as disagreement between techniques remain. Statistical methods that are often utilized in an attempt to extract trends from this data (e.g. locally-weighted scatter plot smoothing, LOWESS) evaluate the mean (with some weighting) of the highly scattered data in a given time interval. However, such methods are only appropriate for estimating the ``true'' value at any given time interval in datasets in which individual data points are being drawn from a single distribution about a mean value. The \pCOtwo proxy compilation shown in Figure \ref{fig:pCO2_proxies} consists of several distinct techniques and samples, each with their own set of assumptions and associated probability distributions that influence their reported values. Therefore, simply taking the mean of these values neglects the fact that some of these techniques/samples are more reliable others. Constraining how robust each \pCOtwo proxy technique is remains an important challenge, and as such the ``true'' \pCOtwo could plausibly lie anywhere within the full range of proxy-based \pCOtwo estimates for any given time interval. Nevertheless, our modeled \pCOtwo values for 15~Ma (Fig. \ref{fig:scenario_pCO2_all}) resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene. Given this scatter in the proxy-based \pCOtwo estimates, we instead infer the Neogene cooling trend from the Miocene benthic foram oxygen isotope record shown in Figure 1 in the main text.

>>>>

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/world_lithology.jpg}
    \caption{Distribution of lithologies at 0.1\degrees $\times$ 0.1\degrees resolution used in GEOCLIM modified from \citet{Hartmann2012a}.}
    \label{fig:world_lithology}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/resolution_sensitivity.pdf}
    \caption{Histograms showing the difference in the area-weighted mean Ca+Mg concentration of the surface in each 0.5\degrees $\times$ 0.5\degrees pixel that results from starting with an 0.1\degrees $\times$ 0.1\degrees GLiM raster versus an 0.05\degrees $\times$ 0.05\degrees GLiM raster. The left panel is calculated over all land pixels, whereas the right panel is calculated over land pixels in the SEAI only. Both panels use the parameter values that produced the highest $r^{2}$ during calibration: metamorphic$_{Ca+Mg}=2500$~mol/m$^{3}$ and sediment$_{Ca+Mg}=2000$~mol/m$^{3}$.}
    \label{fig:resolution_sensitivity}
\end{figure}

We also explore the sensitivity of the area-weighted mean Ca+Mg concentration of the surface in each 0.5\degrees $\times$ 0.5\degrees pixel to the resolution of the underlying GLiM raster (Fig. \ref{fig:resolution_sensitivity}). We find that the difference in the area-weighted mean Ca+Mg concentration of the surface when using the 0.1\degrees $\times$ 0.1\degrees GLiM raster versus the 0.05\degrees $\times$ 0.05\degrees GLiM raster can be well explained by a tight Gaussian distribution ($\sigma$\textless200~mol/m$^{3}$) about a mean of $\sim$0~mol/m$^{3}$. Therefore, using the lower-resolution GLiM raster does not overall bias our results to higher or lower area-weighted mean Ca+Mg concentrations of the surface, and only introduces some noise that has a magnitude that is significantly smaller than the most Ca+Mg-poor lithologic category (i.e. felsics, at 1,521~mol/m$^{3}$).

@@

Proxy-based \pCOtwo estimates for the past 20~m.y. are shown in Figure \ref{fig:pCO2_proxies} \citep{Bereiter2015a, Foster2017a, Ji2018a, Cui2020a}. Even when stringent quality criteria and the latest understanding of each of the \pCOtwo proxies have been applied to available \pCOtwo records, both significant uncertainty in the estimated \pCOtwo for any given data point as well as disagreement between techniques remain. Statistical methods that are often utilized in an attempt to extract trends from this data (e.g. locally-weighted scatter plot smoothing, LOWESS) evaluate the mean (with some weighting) of the highly scattered data in a given time interval. However, such methods are only appropriate for estimating the ``true'' value at any given time interval in datasets in which individual data points are being drawn from a single distribution about a mean value. The \pCOtwo proxy compilation shown in Figure \ref{fig:pCO2_proxies} consists of several distinct techniques and samples, each with their own set of assumptions and associated probability distributions that influence their reported values. Therefore, simply taking the mean of these values neglects the fact that some of these techniques/samples are likely more reliable than the others. Constraining how robust each \pCOtwo proxy technique is remains an important challenge, and as such the ``true'' \pCOtwo could plausibly lie anywhere within the full range of proxy-based \pCOtwo estimates for any given time interval. Nevertheless, our modeled \pCOtwo values for 15~Ma (Fig. \ref{fig:scenario_pCO2_all}) resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene. Given this scatter in the proxy-based \pCOtwo estimates, we instead infer the Neogene cooling trend from the Miocene benthic foram oxygen isotope record shown in Figure 1 in the main text.
